---
title: "Introduction"
author: "Richard Cotton"
date: '`r Sys.Date()`'
output: html_document
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Introduction}
-->

```{r, Setup, echo = FALSE, results = "hide"}
set.seed(19790801)
library(assertive)
knitr::opts_chunk$set(error = FALSE)
```

### *assert* functions

There are times when it is a good idea to check the state of your variables, to
ensure that they have the properties that you think they have.  For example,
if you have a count variable, you might want to check that it is numeric, that
all the values are non-negative, and that all the values are whole numbers.

Base-R has a function called `stopifnot` that lets you perform such checks.

```{r, CountsStopIfNot, error = TRUE, purl = FALSE}
counts <- c(1, 2, 3, 4.5)
stopifnot(
  is.numeric(counts),
  all(counts >= 0),
  isTRUE(all.equal(counts, round(counts)))
)
```

This is OK, but not that easy to read.  Worse, the error messages that it 
produces in the event of failure aren't very user-friendly.

`assertive` provides lots of *assert* functions that provide checks
for specific conditions.  (An *assertion* is software development jargon for a 
check.)  They are designed to make your code easier to read, and to return 
helpful error messages to users in the event of a check failing.

Here's the same example again, written in an `assertive` style.

```{r, CountsAssertive, error = TRUE, purl = FALSE}
counts <- c(1, 2, 3, 4.5)
assert_is_numeric(counts)
assert_all_are_non_negative(counts)
assert_all_are_whole_numbers(counts)
```

Here you see that the error message contains a human readable sentence, followed
by information on the values that caused problems, along with their positions
and reasons for failure.

### *is* and *has* functions

Each of the *assert* functions has an underlying *is* or *has* function.  For
example, `assert_is_numeric` calls `is_numeric`, `assert_all_are_non_negative`
calls `is_non_negative`, and so on.

Some *is* and *has* functions, such as `is_numeric`, return a single logical 
value.

```{r, IsNumeric}
is_numeric(1:6)
is_numeric(letters)
```

When the check passed, `is_numeric` returned `TRUE`, and when it failed, 
`is_numeric` returned `FALSE` with a `cause` attribute explaining the problem.

Where *is* functions return a single value, they have a single corresponding 
*assert* function prefixed by `assert_`.  For example, `is_numeric` is paired 
with `assert_is_numeric`.

Some *is* and *has* functions, such as `is_non_negative`, return a logical 
vector.

```{r, IsNonNegative}
is_non_negative(rnorm(6))
```

`is_non_negative` returned a logical vector which was `TRUE` where the check
passed, and `FALSE` where the check failed.  This time the `cause` attribute
was also vectorised, returning an empty string for the passes and a brief
explanation of the problem for the failures.

Where *is* functions return a vector, there are two corresponding *assert* 
functions, prefixed `assert_all_are_`, and `assert_any_are_`.  For example,
`is_non_negative` is paired with `assert_all_are_non_negative` and 
`assert_any_are_non_negative`.
